---
title: "kate_analysis"
author: "Kate Colvin"
date: "2026-04-12"
output: pdf_document
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(ggplot2)
library(gridExtra)
library(gtsummary)
library(gt)
library(lme4)
library(lmerTest)
library(performance)
library(glmmTMB)
library(DHARMa)

```

1) Did the pilot study provide any evidence for effectiveness of the intervention program (i.e. any clues to answer their questions for primary interests)? Describe and comment on the effect sizes. The primary outcome of interest is SFD (symptom free days in the past two weeks prior to each of the three time points).

--

# i) Formulation of the hypotheses relevant to specific questions and testing those hypotheses

$H_0: $ No difference in the change of symptom-free days (SFD) over time between children receiving Asthma-PASS versus the standard of care. 
$H_A: $ The children receiving Asthma-PASS show a greater improvement in SFD over time compared to the standard of care group. 

$$
H_0: \beta_3 = 0, \quad H_A: \beta_3 > 0
$$

$$
\beta_3 = \text{difference in change of SFD over time between Asthma-PASS (intervention) and control groups}
$$


# ii) Presenting graphical displays and relevant descriptive statistics

```{r}

data <- readxl::read_excel("Proj 4 data.xlsx") %>% 
  janitor::clean_names()

data <- data %>%
  mutate(id = factor(id),
         time = factor(time,
                  levels = c(1, 2, 3),
                  labels = c("Baseline", "6 months", "12 months")),
         group = factor(group,
                   levels = c(0, 1),
                   labels = c("Control", "Intervention")),
         school = factor(school,
                    levels = c(1, 2, 3, 4),
                    labels = c("School 1", "School 2", "School 3", "School 4")))
data

```

Summary Table (Table 1) (Using Chhiring's instead)
```{r, eval = FALSE}

# Nevermind, just using Chhiring's Table 

baseline_data <- data %>%
  filter(time == "Baseline")


table1 <- baseline_data %>%
  select(group, sfd, school) %>%
  tbl_summary(by = group,
              label = list(
                sfd ~ "Symptom-Free Days",
                school ~ "School"),
    digits = all_continuous() ~ 2) %>%
  add_overall() %>%
  bold_labels() %>% 
  modify_caption("**Table 1**: Baseline Symptom-Free Days and School by Intervention Group")

table1

table1_gt <- table1 %>% as_gt()
gtsave(table1_gt, "results/table1.png")


```

Data Visualization

```{r, warning = FALSE}

# Avg. Line Plot

figure1 <- ggplot(data, aes(x = time, y = sfd, color = group, group = group)) +
  stat_summary(fun = mean, geom = "line", size = 1) +
  stat_summary(fun = mean, geom = "point", size = 3) +
  stat_summary(fun.data = mean_cl_normal, geom = "errorbar", width = 0.1) +
  labs(x = "Time", y = "Mean Symptom-Free Days",
       title = "Mean Symptom-Free Days by Group Over Time") +
  theme_minimal() +
  theme(plot.title = element_text(size = 12)) +
  coord_cartesian(ylim = c(6, 14))

# ggsave(filename = "results/line_plot.png", plot = figure1, 
#      width = 8, height = 5, dpi = 300)


# Avg. Line Plot by School

figure3 <- ggplot(data, aes(x = time, y = sfd, group = school, color = school)) +
  stat_summary(fun = mean, geom = "line", size = 1) +
  stat_summary(fun = mean, geom = "point", size = 3) +
  stat_summary(fun.data = mean_cl_normal, geom = "errorbar", width = 0.1) +
  facet_wrap(~ group) +
  labs(x = "Time", y = "Mean Symptom-Free Days", 
       title = "Mean Symptom-Free Days by School and Group Over Time") +
  theme_minimal() +
  theme(plot.title = element_text(size = 12)) +
  coord_cartesian(ylim = c(6, 14))

# ggsave(filename = "results/line_plot_school.png", plot = figure3, 
#      width = 8, height = 5, dpi = 300)


figure1_combined <- grid.arrange(figure1, figure3, nrow = 2)

ggsave(filename = "results/line_plot_combined.png", plot = figure1_combined, 
       width = 6, height = 6, dpi = 300)

```

```{r}

# Boxplot

figure2 <- ggplot(data, aes(x = time, y = sfd, fill = group)) +
  geom_boxplot() +
  # geom_violin(trim = FALSE) +
  scale_y_continuous(limits = c(0, 16)) +
  labs(x = "Time", y = "Symptom-Free Days",
       title = "Distribution of Symptom-Free Days by Time and Group") +
  theme_minimal()

figure2


ggsave(filename = "results/box_plot.png", plot = figure2, 
      width = 8, height = 5, dpi = 300)

```


# iii) Justification and description of the statistical modeling approach used (i.e. write down the statistical models with interpretation of model parameters)

LMM model, interaction term is main variable of interest. 
Included student-level random effect nested by school. 

Tried LMM, poor diagnostics 
Tried Binomial LMM, slightly better diagnostics, but statistically significant over-dispersion
Trying Beta-Binomial (settling on this!) with zero-inflation term

$$
SFD_{ijt} \sim \text{Beta-Binomial}(14,\; p_{ijt},\; \phi)
$$
$$
\text{logit}(p_{ijt}) = \beta_0 + \beta_1 \text{Time}_t + \beta_2 \text{Group}_j + \beta_3 (\text{Time}_t \times \text{Group}_j) + s_j + b_{ij}
$$
$$
s_j \sim N(0, \sigma^2_{\text{school}}, \quad
b_{ij} \sim N(0, \sigma^2_{\text{child}}), \quad
\text{logit}(\pi) = \gamma_0
$$


```{r}

# Plot of zero-inflated data (data distribution) (figure 3)

sfd_dist <- ggplot(data, aes(x = sfd)) +
  geom_histogram(binwidth = 1, fill = "navy") +
  scale_x_continuous(breaks = 0:14) +
  labs(title = "Distribution of Symptom-Free Days",
       x = "SFD (0–14 days)", y = "Count") +
  theme_minimal()

sfd_dist

ggsave(filename = "results/sfd_dist.png", plot = sfd_dist, 
      width = 8, height = 5, dpi = 300)

```

Observations at both extremes (0 and 14 days), indicating potential deviations from beta-binomial.
Zero-inflated approach was added to assess whether there was more zero values than what was expected (beyond what the beta-binomial model could capture). 


# iv) Summary of the results, with appropriate comments, tables and figures to support your answers

-- Beta-Binomial GLMM -- 

```{r}

# Create symptom days
data <- data %>% mutate(failures = 14 - sfd)

# Full Model
bb_glmm_full <- glmmTMB(
  cbind(sfd, failures) ~ time * group + (1 | school/id),
  ziformula = ~ 1, 
  family = betabinomial(link = "logit"),
  data = data)

VarCorr(bb_glmm_full)
# Essentially 0 variation between schools

```

```{r}

model_full_results <- tbl_regression(
  bb_glmm_full,
  label = list(time ~ "Time", 
               group ~ "Group"), 
  conf.int = TRUE,
  conf.level = 0.95) %>%
  bold_labels() %>% 
  bold_p() %>% 
  modify_caption("**Table 2**: Beta-Binomial GLMM Results, Including Treatment by Time Interaction")

model_full_results


table2_gt <- model_full_results %>% as_gt()
gtsave(table2_gt, "results/table2.png")

```

```{r}

# Without interaction
bb_glmm_reduced <- glmmTMB(
  cbind(sfd, failures) ~ time + group + (1 | school/id),
  ziformula = ~ 1, 
  family = betabinomial(link = "logit"),
  data = data)
summary(bb_glmm_reduced)

model_red_results <- tbl_regression(
  bb_glmm_reduced,
  label = list(time ~ "Time", 
               group ~ "Group"), 
  conf.int = TRUE,
  conf.level = 0.95) %>%
  bold_labels() %>% 
  bold_p() %>% 
  modify_caption("**Table 3**: Beta-Binomial GLMM Results, No Treatment by Time Interaction")

model_red_results


table3_gt <- model_red_results %>% as_gt()
gtsave(table3_gt, "results/table3.png")

```

```{r}

# LRT for interaction term
lrt <- anova(bb_glmm_reduced, bb_glmm_full)

lrt_tbl <- as.data.frame(lrt) %>%
  rownames_to_column(var = "Model") %>% 
  select(Model, everything()) %>% 
  rename(df = Df,
         L.Ratio = Chisq,
         `p-value` = `Pr(>Chisq)`) %>%
  mutate(
    logLik = round(logLik, 3),
    AIC = round(AIC, 1),
    BIC = round(BIC, 1),
    L.Ratio = round(L.Ratio, 3),
    `p-value` = signif(`p-value`, 4)) %>% 
  select(Model, df, AIC, BIC, logLik, L.Ratio, `p-value`)

lrt_tbl$Model <- c("Reduced (no interaction)", "Full (with interaction)")

lrt_tbl

gt(lrt_tbl) %>% 
  tab_header(title = "Table 4: Likelihood Ratio Test, inclusion of Treatment by Time Interaction") %>% 
  gtsave("results/table4.png")

```

```{r}

# No overdispersion this time!
check_overdispersion(bb_glmm_reduced)

```

```{r}

# Model diagnostics 

png("results/bb_glmm_diagnostics.png", width = 1200, height = 800)

res <- simulateResiduals(bb_glmm_reduced, n = 1000)
plot(res)

dev.off()



png("results/zi_plot.png", width = 1200, height = 800)

testZeroInflation(res)

dev.off()

```

The residuals vs. fitted plot exhibits a structured sloped-down due to sfd being bounded and discrete (0-14). 
The Q-Q plot indicates some deviation from normality. Overall, the model seems to be an adequate fit to the data.




# Can ignore everything below this! LMM and Binomial GLMM Models 
-> Binomial LMM section has overdispersion test results for model justification!

-- LMM --  

```{r, eval = FALSE}

# Checking school-level clustering term:

# With interaction, school clustering
lmm_full <- lmer(sfd ~ time * group + (1 | school/id),
                   data = data, REML = FALSE)
summary(lmm_full)


# With interaction, no school clustering
lmm_no_school <- lmer(sfd ~ time * group + (1 | id), 
                   data = data, REML = FALSE)
summary(lmm_no_school)


VarCorr(lmm_full)

```

The estimated variance of the school-level random effect was zero, likely due to the small number of clusters (four schools); also, this model had a singular fit. 
Moving forward we exclude school-level random effect, results are identical in the model without the school-specific term. 

```{r, eval = FALSE}

# Interaction, no school clustering
summary(lmm_no_school)

model_int_results <- tbl_regression(
  lmm_no_school,
  label = list(time ~ "Time", 
               group ~ "Group"), 
  conf.int = TRUE,
  conf.level = 0.95) %>%
  bold_labels() %>% 
  bold_p() %>% 
  modify_caption("**Table 2**: LMM Results, Including Treatment by Time Interaction")

model_int_results


table2_gt <- model_int_results %>% as_gt()
gtsave(table2_gt, "results/table2.png")

```

```{r, eval = FALSE}

# No interaction, no school clustering
lmm_reduced <- lmer(sfd ~ time + group + (1 | id),
                      data = data, REML = FALSE)
summary(lmm_reduced)

model_red_results <- tbl_regression(
  lmm_reduced,
  label = list(time ~ "Time", 
               group ~ "Group"), 
  conf.int = TRUE,
  conf.level = 0.95) %>%
  bold_labels() %>% 
  bold_p() %>% 
  modify_caption("**Table 3**: LMM Results, No Treatment by Time Interaction")

model_red_results


table3_gt <- model_red_results %>% as_gt()
gtsave(table3_gt, "results/table3.png")

```

```{r, eval = FALSE}

# LRT
lrt <- anova(lmm_reduced, lmm_no_school)

lrt_tbl <- as.data.frame(lrt) %>%
  rownames_to_column(var = "Model") %>% 
  select(Model, everything()) %>% 
  rename(df = Df,
         L.Ratio = Chisq,
         `p-value` = `Pr(>Chisq)`) %>%
  mutate(
    logLik = round(logLik, 3),
    AIC = round(AIC, 1),
    BIC = round(BIC, 1),
    L.Ratio = round(L.Ratio, 3),
    `p-value` = signif(`p-value`, 4)) %>% 
  select(Model, df, AIC, BIC, logLik, L.Ratio, `p-value`)

lrt_tbl$Model <- c("Reduced (no interaction)", "Full (with interaction)")

lrt_tbl

gt(lrt_tbl) %>% 
  tab_header(title = "Table 4: Likelihood Ratio Test, inclusion of Treatment by Time Interaction") %>% 
  gtsave("results/table4.png")

```

Inclusion of the interaction term does not seem useful, fail to reject null.

Describe and comment on effect size:

In the full model with interaction, the intervention group improved by approximately 1.4–1.6 more symptom-free days than the control group at follow-up (respectively at 12 months and 6 months after baseline). Even though this is a moderate effect size, these results were not statistically significant. In an LRT, this full model was not a statistically significantly better fit than the model without this interaction term. 

Looking at the reduced model, a statistically significant improvement in SFD was seen over time. Being 6 and 12 months past baseline were both associated with a ~2 day increase in SFD, compared to baseline. This is a roughly 15% increase over a 2 week time period. 

Asthma-PASS did not show a statistically significant improvement in SFD compared to the control group. The slight positive direction of the coefficient indicates that the intervention contributes to an improvement in SFD, but the magnitude of this estimate is very small, and is not statistically significant (0.32). 


Model Diagnostics 

```{r, eval = FALSE}

diag_data <- data.frame(
  fitted = fitted(lmm_reduced),
  resid  = resid(lmm_reduced),
  time   = model.frame(lmm_reduced)$time)

# Residuals vs. Fitted values
p1 <- ggplot(diag_data, aes(x = fitted, y = resid)) +
  geom_point(alpha = 0.6) +
  geom_hline(yintercept = 0, color = "red") +
  labs(title = "Residuals vs Fitted",
       x = "Fitted values",
       y = "Residuals")

# QQ-Plot
p2 <- ggplot(diag_data, aes(sample = resid)) +
  stat_qq() +
  stat_qq_line(color = "red") +
  labs(title = "Normal Q-Q Plot")

# Residual Histogram
p3 <- ggplot(diag_data, aes(x = resid)) +
  geom_histogram(bins = 20,
                 fill = "skyblue",
                 color = "black") +
  labs(title = "Histogram of Residuals",
       x = "Residuals",
       y = "Frequency")

# Residuals vs Time
p4 <- ggplot(diag_data, aes(x = time, y = resid)) +
  geom_boxplot() +
  labs(title = "Residuals by Time",
       x = "Time",
       y = "Residuals")

grid.arrange(p1, p2, p3, p4, nrow = 2)

```


Checking that these results are the same for the LMMs with the school-level clustering:

```{r, eval = FALSE}

# With school, with interaction
summary(lmm_full)

# With school, without interaction
lmm_reduced_school <- lmer(sfd ~ time + group + (1 | school/id), 
                   data = data, REML = FALSE)
summary(lmm_reduced_school)

anova(lmm_full, lmm_reduced_school)

```


LMM results are the same (interaction term does not improve fit). 

-- Binomial LMM -- 

```{r, eval = FALSE}

# Create symptom days (failures)
data <- data %>% mutate(failures = 14 - sfd)

# Fit binomial mixed model
glmm_binom <- glmer(
  cbind(sfd, failures) ~ time * group + (1 | school/id),
  data = data,
  family = binomial)
summary(glmm_binom)

check_overdispersion(glmm_binom)

```

```{r, eval = FALSE}

diag_data <- data.frame(
  fitted = fitted(glmm_binom),
  resid  = resid(glmm_binom),
  time   = model.frame(glmm_binom)$time)

# Residuals vs. Fitted values
p1 <- ggplot(diag_data, aes(x = fitted, y = resid)) +
  geom_point(alpha = 0.6) +
  geom_hline(yintercept = 0, color = "red") +
  labs(title = "Residuals vs Fitted",
       x = "Fitted values",
       y = "Residuals")

# QQ-Plot
p2 <- ggplot(diag_data, aes(sample = resid)) +
  stat_qq() +
  stat_qq_line(color = "red") +
  labs(title = "Normal Q-Q Plot")

# Residual Histogram
p3 <- ggplot(diag_data, aes(x = resid)) +
  geom_histogram(bins = 20,
                 fill = "skyblue",
                 color = "black") +
  labs(title = "Histogram of Residuals",
       x = "Residuals",
       y = "Frequency")

# Residuals vs Time
p4 <- ggplot(diag_data, aes(x = time, y = resid)) +
  geom_boxplot() +
  labs(title = "Residuals by Time",
       x = "Time",
       y = "Residuals")

grid.arrange(p1, p2, p3, p4, nrow = 2)

```


Significant overdispersion (dispersion = 1.54, p < 0.001), indicating violation of model assumptions. 
Worse issue than LMM? I think. 
Trying Beta-Binomial now. 


