---
title: "kate_analysis"
author: "Kate Colvin"
date: "2026-04-12"
output: pdf_document
---

TODO:
- Data vis (2-3 plots)
- Effect size

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(ggplot2)
library(gtsummary)
library(gt)
library(lme4)
library(lmerTest)

```

1) Did the pilot study provide any evidence for effectiveness of the intervention program (i.e. any clues to answer their questions for primary interests)? Describe and comment on the effect sizes. The primary outcome of interest is SFD (symptom free days in the past two weeks prior to each of the three time points).

--

# i) Formulation of the hypotheses relevant to specific questions and testing those hypotheses

$H_0: $ No difference in the change of symptom-free days (SFD) over time between children receiving Asthma-PASS versus the standard of care. 
$H_A: $ The children receiving Asthma-PASS show a greater improvement in SFD over time compared to the standard of care group. 

$$
H_0: \beta_3 = 0, \quad H_A: \beta_3 \neq 0
$$

$$
\beta_3 = \text{difference in change of SFD over time between Asthma-PASS (intervention) and control groups}
$$


# ii) Presenting graphical displays and relevant descriptive statistics

```{r}

data <- readxl::read_excel("Proj 4 data.xlsx") %>% 
  janitor::clean_names()

data <- data %>%
  mutate(id = factor(id),
         time = factor(time,
                  levels = c(1, 2, 3),
                  labels = c("Baseline", "6 months", "12 months")),
         group = factor(group,
                   levels = c(0, 1),
                   labels = c("Control", "Intervention")),
         school = factor(school,
                    levels = c(1, 2, 3, 4),
                    labels = c("School 1", "School 2", "School 3", "School 4")))
data

```

Summary Table (Table 1)
```{r}

baseline_data <- data %>%
  filter(time == "Baseline")


table1 <- baseline_data %>%
  select(group, sfd, school) %>%
  tbl_summary(by = group,
              label = list(
                sfd ~ "Symptom-Free Days",
                school ~ "School"),
    digits = all_continuous() ~ 2) %>%
  add_overall() %>%
  bold_labels() %>% 
  modify_caption("**Table 1**: Baseline Symptom-Free Days and School by Intervention Group")

table1

table1_gt <- table1 %>% as_gt()
gtsave(table1_gt, "results/table1.png")


```

Plots
```{r}



```


# iii) Justification and description of the statistical modeling approach used (i.e. write down the statistical models with interpretation of model parameters)

LMM model, interaction term is main variable of interest. 
First, testing the inclusion of random effects by school (am already including random effects by student). 

$$
SFD_{ijt} = \beta_0 + \beta_1 \text{Time}_{t} + \beta_2 \text{Group}_{j} + \beta_3 (\text{Time}_{t} \times \text{Group}_{j}) + s_j + b_{ij} + \epsilon_{ijt}
$$

$$
s_j \sim N(0, \sigma^2_{school}), \quad b_{ij} \sim N(0, \sigma^2_{child}), \quad \epsilon_{ijt} \sim N(0, \sigma^2)
$$


# iv) Summary of the results, with appropriate comments, tables and figures to support your answers

Checking school-level clustering term:
```{r}

# With interaction, school clustering
lmm_full <- lmer(sfd ~ time * group + (1 | school) + (1 | id), 
                   data = data, REML = FALSE)
summary(lmm_full)


# With interaction, no school clustering
lmm_no_school <- lmer(sfd ~ time * group + (1 | id), 
                   data = data, REML = FALSE)
summary(lmm_no_school)


VarCorr(lmm_full)

```

The estimated variance of the school-level random effect was zero, likely due to the small number of clusters (four schools); also, this model had a singular fit. 
Moving forward we exclude school-level random effect, results are identical in the model without the school-specific term. 

```{r}

# Interaction, no school clustering
summary(lmm_no_school)

model_int_results <- tbl_regression(
  lmm_no_school,
  label = list(time ~ "Time", 
               group ~ "Group"), 
  conf.int = TRUE,
  conf.level = 0.95) %>%
  bold_labels() %>% 
  bold_p() %>% 
  modify_caption("**Table 2**: LMM Results, Including Treatment by Time Interaction")

model_int_results


table2_gt <- model_int_results %>% as_gt()
gtsave(table2_gt, "results/table2.png")

```


```{r}

# No interaction, no school clustering
lmm_reduced <- lmer(sfd ~ time + group + (1 | id),
                      data = data, REML = FALSE)
summary(lmm_reduced)

model_red_results <- tbl_regression(
  lmm_reduced,
  label = list(time ~ "Time", 
               group ~ "Group"), 
  conf.int = TRUE,
  conf.level = 0.95) %>%
  bold_labels() %>% 
  bold_p() %>% 
  modify_caption("**Table 3**: LMM Results, No Treatment by Time Interaction")

model_red_results


table3_gt <- model_red_results %>% as_gt()
gtsave(table3_gt, "results/table3.png")

```

```{r}

# LRT
lrt <- anova(lmm_reduced, lmm_no_school)

lrt_tbl <- as.data.frame(lrt) %>%
  rownames_to_column(var = "Model") %>% 
  select(Model, everything()) %>% 
  rename(df = Df,
         L.Ratio = Chisq,
         `p-value` = `Pr(>Chisq)`) %>%
  mutate(
    logLik = round(logLik, 3),
    AIC = round(AIC, 1),
    BIC = round(BIC, 1),
    L.Ratio = round(L.Ratio, 3),
    `p-value` = signif(`p-value`, 4)) %>% 
  select(Model, df, AIC, BIC, logLik, L.Ratio, `p-value`)

lrt_tbl$Model <- c("Reduced (no interaction)", "Full (with interaction)")

lrt_tbl

gt(lrt_tbl) %>% 
  tab_header(title = "Table 4: Likelihood Ratio Test, inclusion of Treatment by Time Interaction") %>% 
  gtsave("results/table4.png")

```

Inclusion of the interaction term does not seem useful, fail to reject null.

Describe and comment on effect size:

In the full model with interaction, the intervention group improved by approximately 1.4–1.6 more symptom-free days than the control group at follow-up (respectively at 12 months and 6 months after baseline). Even though this is a moderate effect size, these results were not statistically significant. In an LRT, this full model was not a statistically significantly better fit than the model without this interaction term. 

Looking at the reduced model, a statistically significant improvement in SFD was seen over time. Being 6 and 12 months past baseline were both associated with a ~2 day increase in SFD, compared to baseline. This is a roughly 15% increase over a 2 week time period. 

Asthma-PASS did not show a statistically significant improvement in SFD compared to the control group. The slight positive direction of the coefficient indicates that the intervention contributes to an improvement in SFD, but the magnitude of this estimate is very small, and is not statistically significant (0.32). 



Model Diagnostics 

```{r}



```





-- 

Checking that these results are the same for the LMMs with the school-level clustering:

Both of these models have a singular fit

```{r}

# With school, with interaction
summary(lmm_full)

# With school, without interaction
lmm_reduced_school <- lmer(sfd ~ time + group + (1 | school) + (1 | id), 
                   data = data, REML = FALSE)
summary(lmm_reduced_school)

```

```{r}

anova(lmm_full, lmm_reduced_school)

```

LMM results are the same (interaction term does not improve fit). 



